---
title: Walleye Pollock in BC case study
author: Sean C. Anderson
---

Load libraries:

```{r}
library(sdmTMB)
library(ggplot2)
theme_set(theme_light())
library(dplyr)
```

Load survey data and survey grid:

```{r}
# surveyjoin::cache_data()
surveyjoin::load_sql_data()

dat <- surveyjoin::get_data("lingcod", regions = "pbs") |>
  mutate(year = lubridate::year(lubridate::ymd(date))) |> 
  select(survey_name, year, lon_start, lat_start, depth_m, effort, catch_weight, common_name)
dat <- sdmTMB::add_utm_columns(dat, ll_names = c("lon_start", "lat_start"), utm_crs = 3156)
table(dat$survey_name, dat$year)
dat <- dat |>
  # Use only complete N/S sampling years
  filter(!(year %in% c(2003, 2004, 2020))) |>
  tidyr::drop_na(effort, catch_weight, depth_m) |>
  # Drop these surveys to be perfectly bienniel
  filter(!(year == 2007 & survey_name == "SYN WCHG")) |>
  filter(!(year == 2021 & survey_name == "SYN WCVI"))
table(dat$survey_name, dat$year)
  
grid <- surveyjoin::dfo_synoptic_grid |>
  sdmTMB::replicate_df("year", unique(dat$year))
grid <- sdmTMB::add_utm_columns(grid, c("lon", "lat"), utm_crs = 3156)

map_data <- rnaturalearth::ne_countries(
  scale = "large",
  returnclass = "sf", country = "canada")
bc_coast <- suppressWarnings(suppressMessages(
  sf::st_crop(map_data,
    c(xmin = -134, ymin = 46, xmax = -120, ymax = 57))))
bc_coast <- sf::st_transform(bc_coast, crs = 3156)
```

```{r}
table(dat$survey_name, dat$year)
```

We can visualize the raw survey data:

```{r raw-data-plot}
.xlim <- range(dat$X * 1000)
.ylim <- range(dat$Y * 1000)
pos_dat <- filter(dat, catch_weight > 0)
zero_dat <- filter(dat, catch_weight == 0)

ggplot() + 
  geom_sf(data = bc_coast) +
  geom_point(data = zero_dat, 
    mapping = aes(X * 1000, Y * 1000), 
    pch = 4, 
    alpha = 0.8, 
    colour = "grey40"
  ) +
  geom_point(data = pos_dat, 
    mapping = aes(X * 1000, Y * 1000, 
      size = catch_weight / effort, 
      colour = catch_weight / effort), 
    pch = 21, 
    alpha = 0.9
  ) +
  facet_wrap(~year) +
  scale_colour_viridis_c(trans = "log10") +
  scale_size_area(max_size = 8) + 
  coord_sf(xlim = .xlim, ylim = .ylim) +
  labs(colour = "Catch (kg) per ha", size = "Catch (kg) per ha") +
  xlab("Longitude (°)") + ylab("Latitude (°)")
```

Set up a finite-element mesh to implement the SPDE approach:

```{r mesh}
mesh <- make_mesh(dat, c("X", "Y"), cutoff = 15)
plot(mesh)
```

We'll initialize an empty list to store our model fits:

```{r fits-list}
fits <- list()
```

First, we'll fit a standard spatiotemporal model with independent year effects, a spatial GMRF, and independent spatiotemporal GMRFs:

```{r fit1}
fits[[1]] <- sdmTMB(
  catch_weight ~ 0 + factor(year),
  data = dat,
  mesh = mesh,
  offset = log(dat$effort),
  family = delta_gamma(type = "poisson-link"),
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  share_range = TRUE,
  anisotropy = TRUE,
  silent = FALSE
)
names(fits)[[1]] <- "IID RF, factor(year)"
```

We can inspect this model with some basic diagnostics:

```{r print}
print(fits[[1]])
```

```{r sanity}
sanity(fits[[1]])
```

```{r anisoplot}
plot_anisotropy(fits[[1]])
```

```{r dharma}
set.seed(1)
simulate(fits[[1]], nsim = 200, type = "mle-mvn") |> 
  dharma_residuals(fits[[1]])
```

We can calculate a relative biomass index from this model fit by predicting from our model onto the survey grid each year and summing predicting density across the domain:

```{r index0}
pred <- predict(fits[[1]], newdata = grid, return_tmb_obj = TRUE)
ind <- get_index(pred)
```

Make a look-up table to keep track of even years for plotting:

```{r}
lu <- data.frame(year = sort(unique(dat$year)))
lu$even <- lu$year %% 2 == 0
lu$survey_group <- ifelse(lu$even, "WCHG + WCVI", "QCS + HS")
```

```{r}
ind |>
  left_join(lu) |> 
  ggplot(aes(year, est, ymin = lwr, ymax = upr)) + 
  geom_ribbon(fill = "grey90") +
  geom_linerange(aes(colour = survey_group)) +
  geom_point(aes(colour = survey_group), size = 2) +
  scale_colour_brewer(palette = "Dark2") +
  ylab("Biomass index") + xlab("Year") + labs(colour = "Survey\ngrouping")

```

Next, we'll fit a sequence of alternative models

```{r}
fits[[2]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "rw"
)
names(fits)[[2]] <- "RW RF"
```

```{r}
fits[[3]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "ar1"
)
names(fits)[[3]] <- "AR1 RF"
```

```{r}
fits[[4]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "rw",
  time_varying = ~ 1,
  time_varying_type = "rw0"
)
names(fits)[[4]] <- "RW RF, RW year"
```

```{r}
fits[[5]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "ar1",
  time_varying = ~ 1,
  time_varying_type = "rw0"
)
names(fits)[[5]] <- "AR1 RF, RW year"
```

```{r}
fits[[6]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "iid",
  time_varying = ~ 1,
  time_varying_type = "rw0"
)
names(fits)[[6]] <- "IID RF, RW year"
```

```{r}
fits[[7]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "off",
  time_varying = ~ 1,
  time_varying_type = "rw0"
)
names(fits)[[7]] <- "Spatial only, RW year"
```

Iterate through our fits to grab the biomass index:

```{r}
future::plan(future::multisession, workers = 4L)
indexes <- furrr::future_map_dfr(fits, \(x) {
  pred <- predict(x, newdata = grid, return_tmb_obj = TRUE)
  get_index(pred)
}, .id = "model", .options = furrr::furrr_options(seed = TRUE))
future::plan(future::sequential)
```

```{r}
saveRDS(indexes, file = "data-generated/bc-indexes-lingcod.rds")
indexes <- readRDS("data-generated/bc-indexes-lingcod.rds")
```


```{r}
indexes |>
  left_join(lu) |> 
  ggplot(aes(year, est, ymin = lwr, ymax = upr)) + 
  geom_ribbon(fill = "grey90") +
  geom_linerange(aes(colour = survey_group)) +
  geom_point(aes(colour = survey_group), size = 2) +
  scale_colour_brewer(palette = "Dark2") +
  facet_wrap(~model) +
  ylab("Biomass index") + xlab("Year") + labs(colour = "Survey\ngrouping")
```
