---
title: Walleye Pollock in BC case study
author: Sean C. Anderson
---

Load libraries:

```{r}
library(sdmTMB)
library(ggplot2)
theme_set(theme_light())
library(dplyr)
```

Load survey data and survey grid:

```{r}
surveyjoin::cache_data()
surveyjoin::load_sql_data()

dat <- surveyjoin::get_data("walleye pollock", regions = "pbs") |>
  mutate(year = lubridate::year(lubridate::ymd(date))) |>
  select(survey_name, year, lon_start, lat_start, depth_m, effort, catch_weight, common_name)
dat <- sdmTMB::add_utm_columns(dat, ll_names = c("lon_start", "lat_start"), utm_crs = 3156)
table(dat$survey_name, dat$year)
dat <- dat |>
  # Use only complete N/S sampling years
  filter(!(year %in% c(2003, 2004, 2020))) |>
  tidyr::drop_na(effort, catch_weight, depth_m) |>
  # Drop these surveys to be perfectly bienniel
  filter(!(year == 2007 & survey_name == "SYN WCHG")) |>
  filter(!(year == 2021 & survey_name == "SYN WCVI"))
table(dat$survey_name, dat$year)

grid <- surveyjoin::dfo_synoptic_grid |>
  sdmTMB::replicate_df("year", unique(dat$year))
grid <- sdmTMB::add_utm_columns(grid, c("lon", "lat"), utm_crs = 3156)

map_data <- rnaturalearth::ne_countries(
  scale = "large",
  returnclass = "sf", country = "canada")
bc_coast <- suppressWarnings(suppressMessages(
  sf::st_crop(map_data,
    c(xmin = -134, ymin = 46, xmax = -120, ymax = 57))))
bc_coast <- sf::st_transform(bc_coast, crs = 3156)
```

```{r}
table(dat$survey_name, dat$year)
```

We can visualize the raw survey data:

```{r raw-data-plot}
.xlim <- range(dat$X * 1000)
.ylim <- range(dat$Y * 1000)
pos_dat <- filter(dat, catch_weight > 0)
zero_dat <- filter(dat, catch_weight == 0)

ggplot() +
  geom_sf(data = bc_coast) +
  geom_point(data = zero_dat,
    mapping = aes(X * 1000, Y * 1000),
    pch = 4,
    alpha = 0.8,
    colour = "grey40"
  ) +
  geom_point(data = pos_dat,
    mapping = aes(X * 1000, Y * 1000,
      size = catch_weight / effort,
      colour = catch_weight / effort),
    pch = 21,
    alpha = 0.9
  ) +
  facet_wrap(~year) +
  scale_colour_viridis_c(trans = "log10") +
  scale_size_area(max_size = 8) +
  coord_sf(xlim = .xlim, ylim = .ylim) +
  labs(colour = "Catch (kg) per ha", size = "Catch (kg) per ha") +
  xlab("Longitude (°)") + ylab("Latitude (°)")
```

Set up a finite-element mesh to implement the SPDE approach:

```{r mesh}
mesh <- make_mesh(dat, c("X", "Y"), cutoff = 15)
plot(mesh)
```

We'll initialize an empty list to store our model fits:

```{r fits-list}
fits <- list()
```

First, we'll fit a standard spatiotemporal model with independent year effects, a spatial GMRF, and independent spatiotemporal GMRFs:

```{r fit1}
fits[[1]] <- sdmTMB(
  catch_weight ~ 0 + factor(year),
  data = dat,
  mesh = mesh,
  offset = log(dat$effort),
  family = delta_gamma(type = "poisson-link"),
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  share_range = TRUE,
  anisotropy = TRUE,
  silent = FALSE
)
names(fits)[[1]] <- "IID RF, factor(year)"
```

We can inspect this model with some basic diagnostics:

```{r print}
print(fits[[1]])
```

```{r sanity}
sanity(fits[[1]])
```

```{r anisoplot}
plot_anisotropy(fits[[1]])
```

```{r dharma}
set.seed(1)
simulate(fits[[1]], nsim = 200, type = "mle-mvn") |>
  dharma_residuals(fits[[1]])
```

We can calculate a relative biomass index from this model fit by predicting from our model onto the survey grid each year and summing predicting density across the domain:

```{r index0}
pred <- predict(fits[[1]], newdata = grid, return_tmb_obj = TRUE)
ind <- get_index(pred)
```

Make a look-up table to keep track of even years for plotting:

```{r}
lu <- data.frame(year = sort(unique(dat$year)))
lu$even <- lu$year %% 2 == 0
lu$survey_group <- ifelse(lu$even, "WCHG + WCVI", "QCS + HS")
```

```{r}
ind |>
  left_join(lu) |>
  ggplot(aes(year, est, ymin = lwr, ymax = upr)) +
  geom_ribbon(fill = "grey90") +
  geom_linerange(aes(colour = survey_group)) +
  geom_point(aes(colour = survey_group), size = 2) +
  scale_colour_brewer(palette = "Dark2") +
  ylab("Biomass index") + xlab("Year") + labs(colour = "Survey\ngrouping")

```

Next, we'll fit a sequence of alternative models

```{r}
fits[[2]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "rw"
)
names(fits)[[2]] <- "RW RF"
```

```{r}
fits[[3]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "ar1"
)
names(fits)[[3]] <- "AR1 RF"
```

```{r}
fits[[4]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "rw",
  time_varying = ~ 1,
  time_varying_type = "rw0",
  priors = sdmTMBpriors(sigma_V = gamma_cv(0.3, 0.5))
)
names(fits)[[4]] <- "RW RF, RW year"
```

```{r}
fits[[5]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "ar1",
  time_varying = ~ 1,
  time_varying_type = "rw0",
  priors = sdmTMBpriors(sigma_V = gamma_cv(0.3, 0.5))
)
names(fits)[[5]] <- "AR1 RF, RW year"
```

```{r}
fits[[6]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "iid",
  time_varying = ~ 1,
  time_varying_type = "rw0",
  priors = sdmTMBpriors(sigma_V = gamma_cv(0.3, 0.5))
)
names(fits)[[6]] <- "IID RF, RW year"
```

```{r}
fits[[7]] <- update(
  fits[[1]],
  formula. = . ~ 1,
  spatiotemporal = "off",
  time_varying = ~ 1,
  time_varying_type = "rw0",
  priors = sdmTMBpriors(sigma_V = gamma_cv(0.3, 0.5))
)
names(fits)[[7]] <- "Spatial only, RW year"
```

Iterate through our fits to grab the biomass index:

```{r}
future::plan(future::multisession, workers = 2L)
indexes <- furrr::future_map_dfr(fits, \(x) {
  pred <- predict(x, newdata = grid, return_tmb_obj = TRUE)
  get_index(pred)
}, .id = "model", .options = furrr::furrr_options(seed = TRUE))
future::plan(future::sequential)
```

```{r}
saveRDS(indexes, file = "data-generated/bc-indexes-lingcod.rds")
indexes <- readRDS("data-generated/bc-indexes-lingcod.rds")
```


```{r}
indexes |>
  left_join(lu) |>
  ggplot(aes(year, est, ymin = lwr, ymax = upr)) +
  geom_ribbon(fill = "grey90") +
  geom_linerange(aes(colour = survey_group)) +
  geom_point(aes(colour = survey_group), size = 2) +
  scale_colour_brewer(palette = "Dark2") +
  facet_wrap(~model) +
  ylab("Biomass index") + xlab("Year") + labs(colour = "Survey\ngrouping")
```

Test for seesaw index on these time series:

```{r}
period2_metric <- function(x, year = seq_along(x), conf = 0.95) {
  stopifnot(
    length(x) == length(year),
    all(x > 0)
  )

  y <- log(x)

  # Alternating survey phases, coded so beta is the phase difference
  phase <- ifelse(year %% 2 == 0, 0.5, -0.5)
  # period2 <- (-1)^year

  fit <- lm(y ~ year + phase)

  beta <- coef(fit)[["phase"]]
  beta_se <- coef(summary(fit))["phase", "Std. Error"]
  ci <- confint(fit, "phase", level = conf)
  A_fun <- \(b) 100 * (exp(abs(b)) - 1)
  A_ci <- A_fun(ci)
  A_bounds <- if (ci[1L] <= 0 && ci[2L] >= 0) {
    c(0, max(A_ci))
  } else {
    range(A_ci)
  }

  c(
    beta = beta,
    beta_se = beta_se,
    beta_lower = ci[1L],
    beta_upper = ci[2L],
    A = A_fun(beta),
    A_lower = A_bounds[1L],
    A_upper = A_bounds[2L]
  )
}

moving_window <- function(x, year, window = 10L, ...) {
  n <- length(x)

  out <- lapply(seq_len(n - window + 1L), function(i) {
    ind <- i:(i + window - 1L)

    res <- period2_metric(
      x = x[ind],
      year = year[ind],
      ...
    )

    data.frame(
      start = year[ind[1L]],
      end = year[ind[length(ind)]],
      t(res)
    )
  })

  do.call(rbind, out)
}
```

```{r}
seesaw_indexes <- indexes |>
  arrange(model, year) |>
  group_by(model) |>
  group_modify(\(.x, .y) {
    moving_window(.x$est, year = .x$year, window = 8L)
  }) |>
  ungroup()
```

```{r}
seesaw_indexes |>
  mutate(year = end) |>
  ggplot(aes(year, exp(beta))) +
  geom_hline(yintercept = 1, lty = 2) +
  geom_linerange(aes(x = year, ymin = exp(beta_lower), ymax = exp(beta_upper))) +
  geom_point() +
  facet_wrap(~model) +
  xlab("Year") +
  ylab("Beta seesaw effect") +
  scale_y_log10()

seesaw_indexes |>
  mutate(year = end) |>
  ggplot(aes(year, A)) +
  geom_linerange(aes(x = year, ymin = A_lower, ymax = A_upper)) +
  geom_point() +
  facet_wrap(~model) +
  xlab("Year") +
  scale_y_continuous(expand = c(0, NA)) +
  ylab("A seesaw effect")

seesaw_indexes |>
  group_by(model) |>
  slice_max(A, n = 1L, with_ties = FALSE) |>
  ungroup() |>
  mutate(model = reorder(model, A)) |>
  ggplot(aes(y = model)) +
  geom_linerange(aes(xmin = A_lower, xmax = A_upper), orientation = "y") +
  geom_point(aes(x = A)) +
  scale_x_continuous(expand = c(0, NA)) +
  xlab("Maximum A seesaw effect") +
  ylab("Model")
```

## Check influence of the random walk prior

```{r}
prior_grid <- tidyr::crossing(
  mean = c(0.1, 0.3, 0.6),
  cv = c(0.2, 0.5, 1)
) |>
  mutate(prior = paste0("mean = ", mean, ", CV = ", cv))

prior_indexes <- purrr::map_dfr(seq_len(nrow(prior_grid)), \(i) {
  fit <- update(
    fits[["IID RF, RW year"]],
    priors = sdmTMBpriors(
      sigma_V = gamma_cv(prior_grid$mean[[i]], prior_grid$cv[[i]])
    )
  )
  pred <- predict(fit, newdata = grid, return_tmb_obj = TRUE)
  get_index(pred) |>
    mutate(prior = prior_grid$prior[[i]])
})

prior_indexes <- prior_indexes |>
  tidyr::extract(
    prior,
    into = c("prior_mean", "prior_cv"),
    regex = "mean = ([0-9.]+), CV = ([0-9.]+)",
    convert = TRUE,
    remove = FALSE
  ) |>
    mutate(prior_mean = paste("Mean =", prior_mean)) |>
    mutate(prior_cv = paste("CV =", prior_cv))

ggplot(prior_indexes, aes(year, est, group = prior)) +
  geom_line(data = filter(prior_indexes, prior == "mean = 0.3, CV = 0.5") |> select(year, est), colour = "black", inherit.aes = FALSE, mapping = aes(x = year, y = est)) +
  geom_line(colour = "red") +
  facet_grid(prior_mean ~ prior_cv) +
  labs(
    x = "Year",
    y = "Abundance index"
  )
```

I think I worked out a nice seesaw metric that works equally for the simulated and real data. This is for walleye pollock BC wide. The below function and then use a moving window. The beta coefficient here is the average difference in log space between the even and odd years after a linear detrending (the 'year' coef). So, exp(beta) gives you the fold difference. E.g., in this plot, 0.5 (same as 2 if flipped), means the odd and even years are on average 2 times different in that window. This example is extreme.


it's nice because you can also apply it to the simulated data by working with the estimated minus true values (residuals) without the detrending part - in fact, it works out to the same thing I was doing with the simulated data, just not in lm() form
because it's fit as a model, you get a CI that helps tell you if you have 'significant' seesawing happening
you could use s(year) if you wanted a non-linear detrending, but for 8-10 years that seems overkill here

I'm also slowly converging on the thinking that we should probably have time_varying intercepts in these models, not just random walk random fields on their own. They seem to do a pretty good job --- they just need a prior to help convergence. Whether the fields should be IID or RW may depend more on the intended purpose.
[4:18 PM]maybe IID fields for things like stock assessment and RW fields for things like SOPO where we are happy to smooth through the pattern and induce more year-to-year correlation
[4:19 PM]I was writing the methods, and writing always makes you think more about these things
[4:20 PM]and as long as the gamma prior on the time-varying SD has a wide enough CV it doesn't seem too influential